Learning Auditory Similarities Between Instrumental Playing Techniques

### 2020-04-04 paper/paper.tex
move up figure 1
finish reworking related work section
move methods section after data section

@@

@@

\subsection*{Timbre modeling as music information retrieval (MIR)}
On the one hand, most publications in music information retrieval cast timbre modeling as an audio classification problem  \cite{martin1998asa,brown1999jasa,eronen2000icassp,herrera2003jnmr,wieczorkowska2003jiis,livshin2004dafx,krishna2004icassp,kaminskyj2005jiis,benetos2006icassp,bhalke2016jiis}.
In this context, the instrumentation of each musical excerpt serves as an unstructured set of ``audio tags,'' encoded as binary outputs within some predefined label space.
Because such tags often belong to the metadata of music releases, the process of curating a training set for musical instrument classification requires little or no human intervention.
Although scraping user-generated content from online music hosting platforms may not always reflect the true instrumentation with perfect accuracy, it offers a scalable and ecologically valid insight onto the acoustic underpinnings of musical timbre.
Furthermore, supplementing user-generated content with the outcome of a crowdsourced annotation campaign allows an explicit verification of instrument tags.
For instance, the Open-MIC dataset \cite{humphrey2018ismir}, maintained by the Community for Open and Sustainable Music Information Research (COSMIR) \cite{mcfee2016ismir}, comprises a vast corpus of 20k polyphonic music recordings spanning 20 instruments as a derivative of the Free Music Archive (FMA) dataset \cite{defferrard2017ismir}.
Another example is the Medley-solos-DB dataset \cite{lostanlen2016ismir}, which comprises 21k monophonic excerpts from 8 instruments as a derivative of the MedleyDB dataset of multitrack music \cite{bittner2014ismir}.
Over the past decade, the availability of large digital audio collections, together with the democratization of high-performance computing on dedicated hardware, has spurred the development of deep learning architectures, and notably deep convolutional networks, in music instrument recognition \cite{mcfee2015ismir,pons2017eusipco,gururani2018ismir}.
Notwithstanding the growing accuracy of these architectures in the large data regime, it remains unclear how to extend them from musical instrument recognition to playing technique recognition, where labeled samples are considerably scarcer.
We refer to \cite{han2017taslp} for a recent review of the state of the art in this domain.

\subsection*{Timbre modeling as music cognition research}
On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations \cite{mcadams2009chapter}.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in deep neural networks.
Nevertheless, while training artificial neurons is prone to statistical overfitting, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex in terms of its spectrotemporal receptive fields (STRFs) \cite{depireux2001jneur}, thus yielding a useful form of inductive bias for downstream machine listening applications.
The STRF of a neuron is a function of time and frequency which represents the optimal predictor of its post-stimulus time histogram (PSTH) during exposure to a diverse range of auditory stimuli \cite{aertsen1981biolcyb}.
The simplest method to compute it in practice is by reverse correlation, i.e.\ by averaging all stimuli that trigger an action potential \cite{deboer1968biomed}.
Historically, STRFs were defined by their Wigner--Ville distribution \cite{flandrin1998book}, thereby sparing the choice of a tradeoff in time--frequency localization, but eliciting cross-term interferences \cite{eggermont1993hearing}.
Since then, the STRF of a neuron was redefined as a spectrographic representation of its spike-triggered average \cite{klein2000compneur}.
Although this new definition is necessarily tied to a choice of spectrogram parameters, it yields more interpretable patterns than a Wigner--Ville distribution.
In particular, a substantial portion of spectrographic STRFs exhibits a ripple-like response around a given region $(t, \lambda)$ of the time--frequency domain \cite{theunissen2000jneur}.
This response can be approximately described by a pair of scalar values: a temporal modulation rate $\alpha$ in Hertz and a frequential modulation rate $\beta$ in cycles per octave.
Interestingly, both $\alpha$ and $\beta$ appear to be arranged in a geometric series and independent from the selected center time $t$ and center frequency $\lambda$.
This observation has led auditory neuroscientists to formulate an idealized computational model for STRF, known as ``full cortical model'' \cite{chi2005jasa}, which densely covers the rate--scale domain $(\alpha, \beta)$.

@@

%\ja{This section is way too long.
%It's important that we situate ourselves with respect to the existing literature, but we don't need to spend 2+ pages doing so.
%This currently reads more like a laundry list of various work related to timbre perception on musical instruments on one hand and a history of STRFs on the other.
%We need to keep in mind what our central themes (wider range of timbre, similarity-based evaluation, simple scattering+linear model) are and link the existing work to them in a concise manner.
%}

@@

@@

\begin{figure}[h!]
\includegraphics[width=0.9\textwidth]{figures/lostanlen2020jasmp_fig1_v3.pdf}
\caption{\csentence{Overview of the proposed approach}}
\label{fig:pipeline}
\end{figure}

@@

Following the above discussed matters, we propose a ranking system that is based on a processing pipeline made of two main components, see Figure \ref{fig:pipeline}.
The first component maps the audio signal of the query and the items of the database into a high-dimensional timbre space using the scattering transform.
The second component warps this timbre space by taking into account perceptual judgments represented as a consensus clustering of the items in the dataset, as presented in the \nameref{sec:survey} section.
This warping, or reweighting, of the space is performed using LMNN.

@@

\subsection*{Expert pre-screening}
Secondly, we reduce the number of IMTs in our study by focusing on those which are deemed to be most relevant.
Here, we define the relevance of an IMT as the possibility of imitating it by means of another IMT from a different instrument.
One example of such imitation is the acoustic similarity between slap tonguing in reed instruments and a snap pizzicato in strings instruments.
To collect perceptual ratings of relevance, we recruited two professors in music composition at the Paris Conservatory (CNSMDP), and explicitly gave them our definition of relevance.
Each of them inspected the entire corpus of $235$ IMTs and annotated them in terms of relevance according to a Likert scale with seven ticks.
In this Likert scale, the value 1 (least relevant) denotes that the IMT under consideration has a timbre that is idiosyncratic of the instrument, and that therefore, it is unlikely that humans will pair it with IMTs from other instruments.
Conversely, the value 7 (most relevant) denotes that the IMT under consideration has a strong similarity with IMTs from one or several other instruments.
Once both experts completed their annotation, we retained all IPTs whose average relevance score was judged equal to 3 or higher.
At the end of this stage of pre-screening, the list of IMT reduces from 235 to $N=78$.
Tables 1 and 2, in the Appendix section of this paper, give a full list of these 78 highly relevant IMTs.
It is worth noting that, according to both experts, the timbre of the accordion was judged too idiosyncratic, regardless of playing technique, to be relevant for this experiment.
Indeed, the accordion is the only instrument in the aforementioned list of 
instrument to have free reeds, keyboard-based actuation, or handheld airflow.
Consequently, regardless of mute and technique, the set of instruments $\mathcal{I}$ in our study contains 15 elements.

Before addressing the question of modeling timbre in the audio domain, we conduct an exploratory study on the structure of similarity judgments themselves.
Specifically, we ask whether human listeners, when confronted with a diverse corpus of IMT samples, tend to cluster those samples in terms of an organological taxonomy.
To answer this question, we derive from the consensus clustering graph $\mathcal{G}_0$ an instrument-wise similarity matrix $\mathbf{A}_{\mathcal{I}}$ whose rows and columns are defined on the finite set $\mathcal{I}$ of instruments.
For every instrumment--instrument pair $(i,j)$, we set the value of $\mathbf{A}_{\mathcal{I}}$ at row $i$ and column $j$ equal to the number of pairs in $\mathcal{G}_0$ from instruments $i$ and $j$ respectively, renormalized by the number of IMT samples from either instrument $i$ or instrument $j$:
\[
\mathbf{A}_{\mathcal{I}}(i,j) = \dfrac{
2\,\Card \big\{ (m, n) \,\vert\, \mathrm{Instrument}(m)=i ; \mathrm{Instrument}(n)=j ; m \overset{\mathcal{G}_0}{\sim} n \big\}
}{
\Card \big\{n \,\vert\, \mathrm{Instrument}(n) \in \{ i, j \} \big\}
}.
\label{eq:instrument-similarity}
\]
%In order to better interpret the information captured by the consensus clustering, we generate two similarity matrices, one for the 15 instruments used, see Figure \ref{fig:consensusVsI}, and one for the 23 PTs used, see Figure \ref{fig:consensusVsPt}. For the instruments similarity matrix, the value within each cell counts the number of times two instruments have been grouped in the consensus clustering, normalized by the average number of IPTs that considered the two instruments. The PTs similarity matrix is computed in the same way.

%$\sigma : I \rightarrow I$ 
%The ordering of the instruments or the PTs on the vertical and horizontal axes is chosen according to the ordering given by an agglomerative hierarchical clustering using the ward criterion of the similarity matrix under inspection. This ordering allows us to interpret the similarity matrices more easily compared to an alphabetical ordering while preserving the  similarity values between each pair of instruments or PTs.

%Interestingly, although several sounds pairs were obtained from the same instrument (but different PTs) and others were obtained using the same PT (but different instruments), the consensus clustering did not favor one of these taxonomies over the other.

%\ml{vl : je pense qu'il faut remplacer la derniere phrase par quelque chose de plus musicologique, en disant que l'on retrouve bien l'organologie, et montrer certains points d'interet sur les pts pour guider le lecteur.}

More generally, even though the block diagonal structure of the matrix $\widetilde{\mathbf{A}}_{\mathcal{T}}$ is less prevalent than in $\widetilde{\mathbf{A}}_{\mathcal{I}}$, playing technique clusters appear to transcend instrument categories.
Nevertheless, these clusters correspond to some basic attributes of qualitative timbre: by and large, the top-left region of the matrix contains sustained sounds whereas the bottom-right region contains percussive sounds.
Interestingly, the ordinary technique (\textt{ord}) appears at the center of the matrix, \ie{} at the intersection between sustained sounds and percussive sounds.
This is because the notion of ``ordinariness'' does not prescribe the same gesture for all instruments: \eg{}, an ordinary guitar sound is expected to be percussive whereas an ordinary flute sound is expected to be sustained.

\section*{\nmu Results}
\label{sec:results}

>>>>

Figure \ref{fig:pipeline} summarizes our experimental protocol: it illustrates how visual annotations (top) can inform feature extraction (center) to produce a nearest-neighbor search engine which is consistent with timbre similarity judgments (bottom).

@@

\begin{figure}[h!]
\includegraphics[width=0.9\textwidth]{figures/lostanlen2020jasmp_fig1_v3.pdf}
\caption{\csentence{Overview of the proposed approach}}
\label{fig:pipeline}
\end{figure}
\clearpage

@@

\subsection*{Timbre modeling as music information retrieval (MIR)}
On the one hand, most publications in music information retrieval cast timbre modeling as an audio classification problem  \cite{martin1998asa,brown1999jasa,eronen2000icassp,herrera2003jnmr,wieczorkowska2003jiis,livshin2004dafx,krishna2004icassp,kaminskyj2005jiis,benetos2006icassp,bhalke2016jiis}.
In this context, the instrumentation of each musical excerpt serves as an unstructured set of ``audio tags,'' encoded as binary outputs within some predefined label space.
Because such tags often belong to the metadata of music releases, the process of curating a training set for musical instrument classification requires little or no human intervention.
Although scraping user-generated content from online music hosting platforms may not always reflect the true instrumentation with perfect accuracy, it offers a scalable and ecologically valid insight onto the acoustic underpinnings of musical timbre.

Furthermore, supplementing user-generated content with the outcome of a crowdsourced annotation campaign allows an explicit verification of instrument tags.
For instance, the Open-MIC dataset \cite{humphrey2018ismir}, maintained by the Community for Open and Sustainable Music Information Research (COSMIR) \cite{mcfee2016ismir}, comprises a vast corpus of 20k polyphonic music recordings spanning 20 instruments as a derivative of the Free Music Archive (FMA) dataset \cite{defferrard2017ismir}.
Another example is the Medley-solos-DB dataset \cite{lostanlen2016ismir}, which comprises 21k monophonic excerpts from 8 instruments as a derivative of the MedleyDB dataset of multitrack music \cite{bittner2014ismir}.

Over the past decade, the availability of large digital audio collections, together with the democratization of high-performance computing on dedicated hardware, has spurred the development of deep learning architectures, and notably deep convolutional networks, in music instrument recognition \cite{mcfee2015ismir,pons2017eusipco,gururani2018ismir}.
Notwithstanding the growing accuracy of these architectures in the large data regime, it remains unclear how to extend them from musical instrument recognition to playing technique recognition, where labeled samples are considerably scarcer.
We refer to \cite{han2017taslp} for a recent review of the state of the art in this domain.

\subsection*{Timbre modeling as music cognition research}
On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations \cite{mcadams2009chapter}.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in deep neural networks.
Nevertheless, while training artificial neurons is prone to statistical overfitting, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex in terms of its spectrotemporal receptive fields (STRFs) \cite{depireux2001jneur}, thus yielding a useful form of inductive bias for downstream machine listening applications.

The STRF of a neuron is a function of time and frequency which represents the optimal predictor of its post-stimulus time histogram (PSTH) during exposure to a diverse range of auditory stimuli \cite{aertsen1981biolcyb}.
The simplest method to compute it in practice is by reverse correlation, i.e.\ by averaging all stimuli that trigger an action potential \cite{deboer1968biomed}.
Historically, STRFs were defined by their Wigner--Ville distribution \cite{flandrin1998book}, thereby sparing the choice of a tradeoff in time--frequency localization, but eliciting cross-term interferences \cite{eggermont1993hearing}.
Since then, the STRF of a neuron was redefined as a spectrographic representation of its spike-triggered average \cite{klein2000compneur}.
Although this new definition is necessarily tied to a choice of spectrogram parameters, it yields more interpretable patterns than a Wigner--Ville distribution.
In particular, a substantial portion of spectrographic STRFs exhibits a ripple-like response around a given region $(t, \lambda)$ of the time--frequency domain \cite{theunissen2000jneur}.
This response can be approximately described by a pair of scalar values: a temporal modulation rate $\alpha$ in Hertz and a frequential modulation rate $\beta$ in cycles per octave.

Interestingly, both $\alpha$ and $\beta$ appear to be arranged in a geometric series and independent from the selected center time $t$ and center frequency $\lambda$.
This observation has led auditory neuroscientists to formulate an idealized computational model for STRF, known as ``full cortical model'' \cite{chi2005jasa}, which densely covers the rate--scale domain $(\alpha, \beta)$.

@@

@@

\subsection*{Expert pre-screening}
Secondly, we reduce the number of IMTs in our study by focusing on those which are deemed to be most relevant.
Here, we define the relevance of an IMT as the possibility of imitating it by means of another IMT from a different instrument.
One example of such imitation is the acoustic similarity between slap tonguing in reed instruments and a snap pizzicato in strings instruments.
To collect perceptual ratings of relevance, we recruited two professors in music composition at the Paris Conservatory (CNSMDP), and explicitly gave them our definition of relevance.
Each of them inspected the entire corpus of $235$ IMTs and annotated them in terms of relevance according to a Likert scale with seven ticks.
In this Likert scale, the value 1 (least relevant) denotes that the IMT under consideration has a timbre that is idiosyncratic of the instrument, and that therefore, it is unlikely that humans will pair it with IMTs from other instruments.
Conversely, the value 7 (most relevant) denotes that the IMT under consideration has a strong similarity with IMTs from one or several other instruments.

Once both experts completed their annotation, we retained all IPTs whose average relevance score was judged equal to 3 or higher.
At the end of this stage of pre-screening, the list of IMT reduces from 235 to $N=78$.
Tables 1 and 2, in the Appendix section of this paper, give a full list of these 78 highly relevant IMTs.
It is worth noting that, according to both experts, the timbre of the accordion was judged too idiosyncratic, regardless of playing technique, to be relevant for this experiment.
Indeed, the accordion is the only instrument in the aforementioned list of 
instrument to have free reeds, keyboard-based actuation, or handheld airflow.
Consequently, regardless of mute and technique, the set of instruments $\mathcal{I}$ in our study contains 15 elements.

Before addressing the question of modeling timbre in the audio domain, we conduct an exploratory study on the structure of similarity judgments themselves.
Specifically, we ask whether human listeners, when confronted with a diverse corpus of IMT samples, tend to cluster those samples in terms of an organological taxonomy.
To answer this question, we derive from the consensus clustering graph $\mathcal{G}_0$ an instrument-wise similarity matrix $\mathbf{A}_{\mathcal{I}}$ whose rows and columns are defined on the finite set $\mathcal{I}$ of instruments.
For every instrumment--instrument pair $(i,j)$, we set the value of $\mathbf{A}_{\mathcal{I}}$ at row $i$ and column $j$ equal to the number of pairs in $\mathcal{G}_0$ from instruments $i$ and $j$ respectively, renormalized by the number of IMT samples from either instrument $i$ or instrument $j$:
\[
\mathbf{A}_{\mathcal{I}}(i,j) = \dfrac{
2\,\Card \big\{ (m, n) \,\vert\, \mathrm{Instrument}(m)=i ; \mathrm{Instrument}(n)=j ; m \overset{\mathcal{G}_0}{\sim} n \big\}
}{
\Card \big\{n \,\vert\, \mathrm{Instrument}(n) \in \{ i, j \} \big\}
}.
\label{eq:instrument-similarity}
\]

More generally, even though the block diagonal structure of the matrix $\widetilde{\mathbf{A}}_{\mathcal{T}}$ is less prevalent than in $\widetilde{\mathbf{A}}_{\mathcal{I}}$, playing technique clusters appear to transcend instrument categories.
Nevertheless, these clusters correspond to some basic attributes of qualitative timbre: by and large, the top-left region of the matrix contains sustained sounds whereas the bottom-right region contains percussive sounds.
Interestingly, the ordinary technique (\texttt{ord}) appears at the center of the matrix, \ie{} at the intersection between sustained sounds and percussive sounds.
This is because the notion of ``ordinariness'' does not prescribe the same gesture for all instruments: \eg{}, an ordinary guitar sound is expected to be percussive whereas an ordinary flute sound is expected to be sustained.

@@

@@

%Following the above discussed matters, we propose a ranking system that is based on a processing pipeline made of two main components, see Figure \ref{fig:pipeline}.
%The first component maps the audio signal of the query and the items of the database into a high-dimensional timbre space using the scattering transform.
%The second component warps this timbre space by taking into account perceptual judgments represented as a consensus clustering of the items in the dataset, as presented in the \nameref{sec:survey} section.
%This warping, or reweighting, of the space is performed using LMNN.

@@

\section*{\nmu Machine listening experiments}
\label{sec:results}
